\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{float}
\usepackage{booktabs} % For improved table formatting
\usepackage{siunitx}  % For number alignment (optional)
\usepackage{subcaption} % For sub-tables


% Tikz package for flowchart
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows}

\tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm, minimum height=1cm,text centered, draw=black, fill=red!30]
\tikzstyle{process} = [rectangle, minimum width=3cm, minimum height=1cm, text centered, draw=black, fill=blue!20]
\tikzstyle{decision} = [diamond, minimum width=3cm, minimum height=1cm, text centered, draw=black, fill=yellow!30]
\tikzstyle{arrow} = [thick,->,>=stealth]

\tikzstyle{block} = [rectangle, minimum width=3.5cm, minimum height=1.2cm, text centered, draw=black, fill=blue!20]
%\tikzstyle{data} = [parallelogram, minimum width=3.5cm, minimum height=1.2cm, text centered, draw=black, fill=green!20]
\tikzstyle{data} = [trapezium, trapezium angle=75, minimum width=3.5cm, minimum height=1.2cm, text centered, draw=black, fill=green!20]
\tikzstyle{command} = [rectangle, minimum width=3cm, minimum height=0.8cm, text centered, draw=black, fill=green!20]

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{ECG Classification}
\author{\IEEEauthorblockN{
    %\IEEEauthorblockN{1\textsuperscript{st} Lewis G. Richter}
    \IEEEauthorblockN{Lewis G. Richter}
    \IEEEauthorblockA{\textit{AI \& Automation} \\
    \textit{University West}\\
    Trollhättan, Sweden \\
    lewis.richter@student.hv.se}
}}

\maketitle

\begin{abstract}
Insert abstract here.
\end{abstract}

\begin{IEEEkeywords}
MIT-BIH Arrhythmia Dataset,Heartbeat Classification, ECG Heartbeat Classification, ECG Heartbeat Detection, ECG Heartbeat Analysis, ECG Heartbeat Recognition, ECG Heartbeat Classification using Deep Learning, ECG Heartbeat Classification using Convolutional Neural Networks, ECG Heartbeat Classification using Long Short-Term Memory Networks, ECG Heartbeat Classification using Gated Recurrent Units
\end{IEEEkeywords}

\section{Administrative Details}
The project is a part of the course ``Deep Learning'' in the masters degree of ``Artificial Intelligence and Automation'' at University West, Trollhättan, Sweden. The project is supervised by Dr. Amit Kumar. The project is carried out by Lewis G. Richter.

The project files are available on \textit{GitHub} at \textit{https://github.com/RichLewi/ECGClassification.git}.

The project is effected by the EU Artificial Intelligence Regulations (EU) \cite{b2} as it regulates AI applications based on a risk-based approach. Since Electrocardiogram (ECG) classification involves medical data and health diagnosis, it may fall under the category of high-risk AI systems due to its potential impact on people's health and safety. Compliance with the regulations requires ensuring data privacy and compliance with the General Data Protection Regulation (GDPR), ensuring transparency and traceability, and ensuring human oversight and accountability to avoid bias and discrimination. Before deploying the AI system, it is necessary to conduct a risk assessment and ensure that the system follows medical certification standards. 

\input{sections/Requirement-DataAnalysis.tex}

\input{sections/SystemEngineering.tex}

\input{sections/AlgorithmDesign.tex}

\input{sections/ConsolidationOfPerformance-FutureWork.tex}

\input{sections/Conclusion.tex}

\bibliographystyle{IEEEtran}
\bibliography{Insert your bibliography file here}

\begin{thebibliography}{00}
    %\bibitem{citekey} Author(s), “Title of the Paper,” Journal Name, vol. xx, no. xx, pp. xxx-xxx, Month Year.
    \bibitem{b1} Shayan Fazeli, "ECG Heartbeat Categorization Dataset."  https://www.kaggle.com/datasets/shayanfazeli/heartbeat (accessed Mar. 11, 2025).

    \bibitem{b2} “Regulation - EU - 2024/1689 - EN - EUR-LEX.” https://eur-lex.europa.eu/legal-content/EN/TXT/?uri=CELEX\%3A32024R1689 (accessed Mar. 11, 2025).

    \bibitem{b3} S. Raghav and A. K. Mishra, "Fractal feature based ECG arrhythmia classification," TENCON 2008 - 2008 IEEE Region 10 Conference, Hyderabad, India, 2008, pp. 1-5, doi: 10.1109/TENCON.2008.4766475. keywords: {Fractals;Electrocardiography;Databases;Heart rate variability;Frequency;Testing;Classification algorithms;Pattern recognition;Heart beat;Signal processing algorithms;Beat classification;ECG arrhythmia;Fractal dimension},

    \bibitem{b4} Youhe Huang and Hongru Li and Xia Yu, "A novel time representation input based on deep learning for ECG classification," Biomedical Signal Processing and Control, vol. 83, p. 104628, 2023, doi: https://doi.org/10.1016/j.bspc.2023.104628. keywords: {Heartbeat classification, Time characteristic representation, Information Fusion, Deep learning}

    \bibitem{b5} Y. Ansari, O. Mourad, K. Qaraqe, and E. Serpedin, “Deep Learning for ECG Arrhythmia Detection and Classification: An Overview of Progress for Period 2017–2023,” Frontiers in Physiology, vol. 14, doi: 10.3389/fphys.2023.1246746.
\end{thebibliography}

\end{document}